Causal.jl:\\A Modeling and Simulation Framework for Causal Models

### 2021-07-03 paper/paper.tex
Added white space before left paranthesis.

In this study, Causal.jl, a modeling and simulation framework for causal models, is introduced \cite{causal}. The aim is to model large scale complex system networks easily and to provide fast and effective simulations. For this purpose, Julia, an open-source, high level, general purpose dynamical programming language designed for high-performance numerical analysis and computational science, has been used. Although Julia is a dynamical language, owing to its Just-in-Time(JIT) compiler developed on Low Level Virtual Machine(LLVM), it can reach the high-speed performance of static languages such as C\cite{bezanson2017julia,julialang}. It supports various parallel computing techniques at thread and process levels. In addition to Julia's standard library, numerous specialized packages developed for different fields such as data science, scientific computing, are also available. Julia's high-speed performance and parallel computing support are essential in meeting the need to design a  fast and effective simulation environment. Julia's syntax can be enlarged purposefully using its metaprogramming support. The analyzes scope of the simulation framework can be extended with new plugins that can be easily defined. It is possible to analyze discrete or continuous-time, static, or dynamical systems. In particular, it is possible to simulate dynamical systems modeled by ordinary, random ordinary, stochastic, delay differential, differential-algebraic, and/or discrete difference equations simultaneously. Unlike its counterparts, the models do not evolve at once for the whole simulation duration. Instead, the model components evolve between sampling time intervals individually. While individual evolution of the components enables the simulation of systems consisting of components represented by different mathematical models, parallel evolution of components increases the simulation performance.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/ComponentsCompressed/components_compresses.pdf}
    \caption{Component types in Causal.jl. $u, x, y$ signify the input, state, output of a system, respectively. $t$ and $k$ are continuous and discrete time variable, respectively. $W$ is the stochastic process corresponding to the noise and $x_\tau$ is the state $x$ delayed by $\tau$ seconds in time. $f$(and $h$) is the right-hand-side function of the differential equation representing the system and $g$ is the readout function.}
    \label{fig: component types}
\end{figure}

A static system is described by a readout equation solely. When triggered, a static system reads its input data from its input port, calculates its output according to its readout function, and writes the result to its output port. In dynamic systems, however, the system behavior is characterized by the states. The output of a dynamical system depends on its input, previous state(s), and time. Therefore, a dynamical system is described by a differential equation and a readout equation, classically. When triggered, a dynamical system reads its input from its input port, updates its state according to the differential equation, calculates its output according to its readout function, and writes the result to its output port. Being developed on top of DifferentialEquations.jl, Causal.jl is capable of simulating the dynamical systems represented by differential equations in the form of the ordinary differential equations(ODE), differential-algebraic equations(DAE), random ordinary differential equations(RODE), stochastic differential equations(SDE), delay differential equations(DDE) or discrete difference equations \cite{rackauckas2017differentialequations}. Most of the available simulation environments allow the systems represented by ordinary differential equations or differential-algebraic equations\cite{elmqvist1978structured,nytsch2006advanced,zimmer2008introducing,mosterman2002hybrsim,van2001variables,giorgidze2009higher,pfeiffer2012pysimulator,simulink}. Therefore, analyzes such as noise or delay analysis or unexpected change of system parameters cannot be performed in these simulation environments, easily. On the contrary, Causal.jl makes it possible for all these analyses to be performed owing to its ability to solve such a wide range of differential equations.

@@

The data transferred to a port are transferred to its connections. The data transfer through the connections is performed over the links of the connections. The links are built on top of unbuffered Julia channels. The data written to(read from) a link is written to(read from) its channel. Running Julia tasks bound to the channels must exist for the data to flow through these channels. Julia tasks are the control flow features that allow calculations to be flexibly suspended and maintained without directly communicating the task scheduler of the operating system\cite{julialang}. The communication and the data exchange between the tasks are carried out through Julia's channels to which they are bound.

@@

In Causal.jl, it is possible to simulate models that include components represented by different types of equations. In Figure \ref{fig: coupled model} is given such an example model. \texttt{odeds} is a dynamical system represented by the ordinary differential equation in (\ref{eq: coupled ode}) where where $x_1, y_1, z_1$ are the state variables and $a=35, b=3, c = 28$ are system parameters.
\begin{equation}
    \begin{split}
        \dot{x}_1 &= a (y_1 - x_1) \\
        \dot{y}_1 &= (c - 1) x_1 + c y_1 - x_1 z_1 \\
        \dot{z}_1 &= x_1 y_1 - b z_1
    \end{split}
    \label{eq: coupled ode}
\end{equation}
\texttt{sdeds} is dynamical system that is represented by the stochastic differential equation in (\ref{eq: coupled sde}) where $x_2, y_2, z_2$ are the state variables, $u=[u_1,u_2, u_3]$ is the input, $\eta$ is noise strength, $W_t$ is the Wiener process corresponding to the noise and $a, b, c$ are the same as those in (\ref{eq: coupled ode}).
\begin{equation}
    \begin{split}
        dx_2 &= (a (y_1 - x_1) + u_1) dt + \eta dW_t \\
        dy_2 &= ((c - 1) x_1 + c y_1 - x_1 z_1)dt + \eta dW_t \\
        dz_2 &= (x_1 y_1 - b z_1 + u_3) dt + \eta dW_t 
    \end{split}
    \label{eq: coupled sde}
\end{equation}
\texttt{gain} is a static system whose input output relation is given in (\ref{eq: gain}) where $u, y, \epsilon$ are the input, output and gain coefficient of the system. 
\begin{equation}
    y = \epsilon u
    \label{eq: gain}
\end{equation}
\texttt{writer1} and \texttt{writer2} are used to record the outputs of \texttt{odeds} and \texttt{sdeds}, respectively, while \texttt{writer3} is used to record the fast Fourier transform(FFT) of the data flowing out of \texttt{odeds}\cite{proakis2004digital}.

>>>>

In this study, Causal.jl, a modeling and simulation framework for causal models, is introduced \cite{causal}. The aim is to model large scale complex system networks easily and to provide fast and effective simulations. For this purpose, Julia, an open-source, high level, general purpose dynamical programming language designed for high-performance numerical analysis and computational science, has been used. Although Julia is a dynamical language, owing to its Just-in-Time (JIT) compiler developed on Low Level Virtual Machine (LLVM), it can reach the high-speed performance of static languages such as C\cite{bezanson2017julia,julialang}. It supports various parallel computing techniques at thread and process levels. In addition to Julia's standard library, numerous specialized packages developed for different fields such as data science, scientific computing, are also available. Julia's high-speed performance and parallel computing support are essential in meeting the need to design a  fast and effective simulation environment. Julia's syntax can be enlarged purposefully using its metaprogramming support. The analyzes scope of the simulation framework can be extended with new plugins that can be easily defined. It is possible to analyze discrete or continuous-time, static, or dynamical systems. In particular, it is possible to simulate dynamical systems modeled by ordinary, random ordinary, stochastic, delay differential, differential-algebraic, and/or discrete difference equations simultaneously. Unlike its counterparts, the models do not evolve at once for the whole simulation duration. Instead, the model components evolve between sampling time intervals individually. While individual evolution of the components enables the simulation of systems consisting of components represented by different mathematical models, parallel evolution of components increases the simulation performance.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/ComponentsCompressed/components_compresses.pdf}
    \caption{Component types in Causal.jl. $u, x, y$ signify the input, state, output of a system, respectively. $t$ and $k$ are continuous and discrete time variable, respectively. $W$ is the stochastic process corresponding to the noise and $x_\tau$ is the state $x$ delayed by $\tau$ seconds in time. $f$ (and $h$) is the right-hand-side function of the differential equation representing the system and $g$ is the readout function.}
    \label{fig: component types}
\end{figure}

A static system is described by a readout equation solely. When triggered, a static system reads its input data from its input port, calculates its output according to its readout function, and writes the result to its output port. In dynamic systems, however, the system behavior is characterized by the states. The output of a dynamical system depends on its input, previous state(s), and time. Therefore, a dynamical system is described by a differential equation and a readout equation, classically. When triggered, a dynamical system reads its input from its input port, updates its state according to the differential equation, calculates its output according to its readout function, and writes the result to its output port. Being developed on top of DifferentialEquations.jl, Causal.jl is capable of simulating the dynamical systems represented by differential equations in the form of the ordinary differential equations (ODE), differential-algebraic equations (DAE), random ordinary differential equations (RODE), stochastic differential equations (SDE), delay differential equations (DDE) or discrete difference equations \cite{rackauckas2017differentialequations}. Most of the available simulation environments allow the systems represented by ordinary differential equations or differential-algebraic equations\cite{elmqvist1978structured,nytsch2006advanced,zimmer2008introducing,mosterman2002hybrsim,van2001variables,giorgidze2009higher,pfeiffer2012pysimulator,simulink}. Therefore, analyzes such as noise or delay analysis or unexpected change of system parameters cannot be performed in these simulation environments, easily. On the contrary, Causal.jl makes it possible for all these analyses to be performed owing to its ability to solve such a wide range of differential equations.

@@

The data transferred to a port are transferred to its connections. The data transfer through the connections is performed over the links of the connections. The links are built on top of unbuffered Julia channels. The data written to (read from) a link is written to (read from) its channel. Running Julia tasks bound to the channels must exist for the data to flow through these channels. Julia tasks are the control flow features that allow calculations to be flexibly suspended and maintained without directly communicating the task scheduler of the operating system\cite{julialang}. The communication and the data exchange between the tasks are carried out through Julia's channels to which they are bound.

@@

In Causal.jl, it is possible to simulate models that include components represented by different types of equations. In Figure \ref{fig: coupled model} is given such an example model. \texttt{odeds} is a dynamical system represented by the ordinary differential equation in (\ref{eq: coupled ode}) where where $x_1, y_1, z_1$ are the state variables and $a=35, b=3, c = 28$ are system parameters.
\begin{equation}
    \begin{split}
        \dot{x}_1 &= a (y_1 - x_1) \\
        \dot{y}_1 &= (c - 1) x_1 + c y_1 - x_1 z_1 \\
        \dot{z}_1 &= x_1 y_1 - b z_1
    \end{split}
    \label{eq: coupled ode}
\end{equation}
\texttt{sdeds} is dynamical system that is represented by the stochastic differential equation in (\ref{eq: coupled sde}) where $x_2, y_2, z_2$ are the state variables, $u=[u_1,u_2, u_3]$ is the input, $\eta$ is noise strength, $W_t$ is the Wiener process corresponding to the noise and $a, b, c$ are the same as those in (\ref{eq: coupled ode}).
\begin{equation}
    \begin{split}
        dx_2 &= (a (y_1 - x_1) + u_1) dt + \eta dW_t \\
        dy_2 &= ((c - 1) x_1 + c y_1 - x_1 z_1)dt + \eta dW_t \\
        dz_2 &= (x_1 y_1 - b z_1 + u_3) dt + \eta dW_t 
    \end{split}
    \label{eq: coupled sde}
\end{equation}
\texttt{gain} is a static system whose input output relation is given in (\ref{eq: gain}) where $u, y, \epsilon$ are the input, output and gain coefficient of the system. 
\begin{equation}
    y = \epsilon u
    \label{eq: gain}
\end{equation}
\texttt{writer1} and \texttt{writer2} are used to record the outputs of \texttt{odeds} and \texttt{sdeds}, respectively, while \texttt{writer3} is used to record the fast Fourier transform (FFT) of the data flowing out of \texttt{odeds}\cite{proakis2004digital}.
